---
title: "Navigating Data Types in R"
author: "Lindsay N. Hayes"
date: 2025-07-03
draft: false
editor_options: 
  chunk_output_type: console
---

## About the activity

1)  Access the Quarto document [here](https://github.com/lindsaynhayes/introR2025/blob/main/activities/classwork02/classwork_L02.qmd).

2)  Download the raw file.

3) Open it in RStudio.

We will work our way through this quarto document together during class. The activity will cover using R as a calculator, creating R objects, and exploring the features of a data set.

## First Load the Tidyverse Package

```{r}
#| label: packages
library(tidyverse)
```

## Ways to Use R

### 1. Arithmetic

```{r}
# example: addition/subtraction/multiplication/division

193 + 45
2050 - 2025
50/250 * 100

# activity: assign the variable my_age as the current year minus your year of birth. 

```

### 2. Create R objects

```{r}
# vector
# c(..., recursive = FALSE, use.names = TRUE)
x <- c(1:225)
class(x)

# matrix
# matrix(data, nrow, ncol, byrow, dimnames)
y <- matrix(1:225, nrow=15, ncol=15, byrow = FALSE)
class(y)

# logical
# testing each variable in the vector and outputting TRUE or FALSE
over100 <- x > 100
table(over100)
```

### Your Turn

### A. Create a vector of all the `homeworlds` in starwars using the `starwars` data.

```{r}

# create vector called "homeworlds" and assign it the value "homeworld" from the starwars data set

glimpse(starwars)

homeworlds <- starwars$homeworld

homeworlds

# how many worlds are there? hint: use the unique function

unique(homeworlds)

starwars |> count(homeworld) # NOTE IT INCLUDES missing values (NA)

# is there a world called "Ohio"? how would you test this with code?

starwars |> count(homeworld == "Ohio")

table(homeworlds == "Ohio")


# How many characters live on Naboo?

table(homeworlds == "Naboo")
starwars |> filter(homeworld == "Naboo") |> count(homeworld)

# Who lives on Naboo? (hint use the "names" variable in the starwars data and the "which" function)

starwars |> filter(homeworld == "Naboo") |> select(name)

homeworlds == "Naboo"
which(homeworlds == "Naboo")
starwars$name[which(homeworlds == "Naboo")]
```

### B. Import the `taylor` package and explore the `taylor_album_songs` dataframe

```{r}

# Try loading in the taylor package and viewing the taylor_album_songs dataframe

# install.packages("taylor")

library(taylor)
taylor <- taylor_album_songs

# what is the "class" of the object taylor?



# what is the "class" of the object taylor_album_songs?


# what types of data are in the object taylor_album_songs?


# change the "key_mode" from class "character" to class "factor"


# How many albums are in the data set & how many songs on each album?


```


### C. Which numeric song features are correlated with one another? Hint create a correlation matrix.

```{r}
# look at the columns in taylor_album_songs and pick what features of the data you want to explore (hint choose all the numeric variables)

# create a matrix of those features

# pipe the matrix into GGally::ggpairs(), evaluate the correlation matrix

# Which values show a positive correlation? Which values show a negative correlation?

#var1 <-
#var2 <-
#var3 <- 
#var4 <- 


# Plot the correlations

#ggplot(taylor_album_songs, aes(x=var1, y=var2)) + geom_point(size=3)

#ggplot(taylor_album_songs, aes(x=var3, y=var4)) + geom_point(size=3)


```