---
title: "Iterations and Conditional Statements"
author: "Lindsay N. Hayes"
date: 2026-07-17
draft: false
editor_options: 
  chunk_output_type: console
---
## About the activity

1)  Access the Quarto document.

2)  Download the quarto file and data files. [here](https://github.com/lindsaynhayes/introR2026/blob/main/activities/classwork06/)

3) Open it in RStudio.

4) Check your working directory!

We will work our way through this quarto document together during class. The activity will cover writing and evaluating different types of iterative loops.

## Load the Packages
The `tidyverse` library includes `ggplot2` and I added `cowplot` to make the plots prettier. We will use data provided on GitHub rather than a package. 
```{r}
#| label: packages
#| message: false

library(tidyverse)
library(cowplot)
```


## Looping Plots
```{r}

## Load Sample Data
norm.counts <- read.csv("norm.counts.csv", row.names = 1)
norm.counts[1:5,1:6]

## Generate Plot Loop
goi <- c("Dat")

for (i in goi){
pgene = which(row.names(norm.counts) == i)
exp <- as.numeric(t(norm.counts[pgene,]))
group <- as.character(rep(c("CON", "EXP"), each = 24))
df <- data.frame(group, exp)

p <- ggplot(df, aes(x = group, y = exp, fill = group)) + geom_boxplot() + geom_point() + 
  scale_fill_manual(values = c("cornflowerblue", "magenta3")) +
  theme_cowplot() +
  labs(subtitle = i) + ylab("normalized expression")

print(p)
#ggsave(filename=paste("plots/", i,".jpeg", sep = ""), width = 80, height = 80, units = "mm")
}

```

### try it with some other genes of your interest, try saving the outputs
```{r}



```



## Read in a multiple files
This kind of loop is useful when you generated hundreds of files that need to be pooled together into one master data sheet. Such as in image analysis with ImageJ or Imaris. 
```{r}
total_files=list.files("data_files/")
total_files

df<-read.csv(paste0("data_files/", total_files[1]), header = FALSE)
head(df)

for(i in 2:length(total_files)){
  tmp <- read.csv(paste0("data_files/", total_files[i]), header=FALSE)
  df <- rbind(df, tmp)
  print(i)
}
colnames(df) <- c("filename","no.intersections","radius.um","x")
dim(df)
# write.csv(df,"data/Sholl.csv")
```

```{r}
## Tidyverse Version

paths <- list.files("data_files", full.names = TRUE)
paths

# \(x) is called an anonymous function
df2 <- paths |> map(\(path) read_csv(path, col_names = FALSE)) |> list_rbind() 

dim(df2)
## this is wildly tidy, just 3 lines!
```


## mutate to assign DEGs
In the next example, we have 2 sets of differentially expressed genes between treated and untreated for example. We want to see what our control and experimental groups have in common or differences in their response to treatment.
```{r}
# first read in the data
con <- read.csv("deg_con.csv", row.names = 1)
exp <- read.csv("deg_exp.csv", row.names = 1)

# this is typical data you get from a differential expression analysis with fold change, stat, pvalue, and adjusted p-value. 

# Lets combine the two dataframes
all.equal(rownames(con), rownames(exp)) # first check that the rows are properly aligned

df <- cbind(con, exp) # combine both analyses into one dataframe, use inner_join() if the rows don't match perfectly.
df$gene.name <- rownames(df)
#rename the columns to be more useful
colnames(df) <- c("baseMean.CON","log2FC.CON","lfcSE.CON","stat.CON","pval.CON","padj.CON","baseMean.EXP","log2FC.EXP","lfcSE.EXP","stat.EXP","pval.EXP","padj.EXP", "gene.name")

# lets take a look before we create groups

ggplot(df, aes(x = log2FC.CON, log2FC.EXP, color = padj.CON)) + geom_point() + theme_cowplot()

# now we are ready to create our groups!
glimpse(df)

tmp <- df |> mutate(cat = case_when(
                   padj.CON < 0.05 & padj.EXP > 0.05 ~ "CON only",
                   padj.CON > 0.05 & padj.EXP < 0.05 ~ "EXP only",
                   padj.CON < 0.05 & padj.EXP < 0.05 ~ "both",
                   .default = "neither"))
table(tmp$cat)

# There are NAs generated in the data analysis so this can be inaccurate we can write a more careful version

df <- df |> mutate(CON.sig = ifelse(padj.CON < 0.05 & !is.na(padj.CON), "YES", "NO"),
                   EXP.sig = ifelse(padj.EXP < 0.05 & !is.na(padj.EXP), "YES", "NO"),
                   cat = case_when(CON.sig == "YES" & EXP.sig == "NO" ~ "CON only",
                                    CON.sig == "NO" & EXP.sig == "YES" ~ "EXP only",
                                    CON.sig == "YES" & EXP.sig == "YES" ~ "both",
                                    .default = "neither"))

table(df$cat)

# lets plot it now that we have the group
ggplot(df, aes(x = log2FC.CON, log2FC.EXP, color = cat)) + geom_point() + theme_cowplot()

# the neither is in the way lets clear it out
df |> filter(cat != "neither") |>
  ggplot(aes(x = log2FC.CON, log2FC.EXP, color = cat)) + geom_point() + theme_cowplot() +
  geom_hline(yintercept = 5) +
  geom_vline(xintercept = 5)

```


### Now write a loop to identify the significant genes that are up-regulated > 5x fold change and DE in either both groups, only control, or only exp

```{r}

# tidy version
df |> mutate(keep = log2FC.CON >5 | log2FC.EXP>5) |> count(keep, cat)

# most right version
df |> mutate(Up5xsig = case_when(cat == "CON only" & log2FC.CON > 5  ~ "UP in CON, sig CON",
                                 cat == "EXP only" & log2FC.EXP > 5  ~ "UP in EXP, sig EXP",
                                 cat == "both"     & log2FC.CON > 5  ~ "UP in CON, DE both",
                                 cat == "both"     & log2FC.EXP > 5  ~ "UP in EXP, DE both",
                                 .default = "neither")) |>
      count(Up5xsig)


df<-df |> mutate(Up5xsig = case_when(cat == "CON only" & log2FC.CON > 5  ~ "UP in CON, sig CON",
                                 cat == "EXP only" & log2FC.EXP > 5  ~ "UP in EXP, sig EXP",
                                 cat == "both"     & log2FC.CON > 5  ~ "UP in CON, DE both",
                                 cat == "both"     & log2FC.EXP > 5  ~ "UP in EXP, DE both",
                                 .default = "neither"))

df$gene.name[which(df$Up5xsig != "neither")]
```





