---
title: "Data Wrangling with tidyverse"
author: "Lindsay N. Hayes"
date: 2026-07-13
draft: false
editor_options: 
  chunk_output_type: console
---

## About the activity

1)  Access the Quarto document [here](https://github.com/lindsaynhayes/introR2026/blob/main/activities/classwork04/classwork_L04.qmd).

2)  Download the raw file.

3) Open it in RStudio.

We will work our way through this quarto document together during class. The activity will cover reshaping, filtering, and summarizing data using tidyverse principles.

## Load the Tidyverse Package
```{r}
#| label: packages
#| message: false

library(tidyverse)
library(cowplot)
```


# Reshaping and Summarizing Data
A common type of data that requires reshaping is time course data. 

Using tidyverse principles answer the questions below:

## 1. On average which month had the most and least passengers in the `AirPassengers` data?

The `AirPassengers` data which is a time-series of data representing the monthly international airline passenger numbers from January 1949 to December 1960. Search for AirPassengers in the `Help` to learn more about the dataset. 

```{r}
# Load and inspect the data, a little reshaping here to get in to an easy to read format for you.

AP_matrix <- matrix(AirPassengers, nrow = length(unique(floor(time(AirPassengers)))), byrow = TRUE)
colnames(AP_matrix) <- month.abb
rownames(AP_matrix) <- unique(floor(time(AirPassengers)))
AP_df <- as.data.frame(AP_matrix)
AP_df$Year <- rownames(AP_matrix)
rm(AP_matrix)
```

### A. Is the data long or wide? What form does it need to be in? How can you convert to the form you need?
```{r}
AP_df

AP_long <- AP_df |> pivot_longer(cols = 1:12, names_to = "months", values_to = "no.passengers")

```

### B. How can we extract the the most and least traveled months each year? 
```{r}

AP_long |> 
  group_by(months) |>
  summarize(avg = mean(no.passengers)) |>
    arrange(avg)

AP_long |> 
  ggplot(aes(x = months, y = no.passengers, fill = months)) + 
  geom_boxplot() + geom_point() +
  theme_cowplot()
  
AP_long$months <- factor(AP_long$months, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

AP_long |> 
  ggplot(aes(x = months, y = no.passengers, fill = months)) + 
  geom_boxplot() + geom_point(size = 1) +
  theme_cowplot() +
  ylim(0, 650)

```

## 2. What was the percent increase in passengers each year between Aug and Nov?
```{r}
AP_long

AP_df |> mutate(ratio = ((Aug/Nov)-1) *100) |>
  select(ratio, Year)

AP_df |> mutate(ratio = ((Aug/Nov)-1) *100) |>
  select(ratio, Year) |>
  ggplot(aes(x = Year, y = ratio)) + geom_col() + theme_cowplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

```


## 3. Which diet lead to heavier chicks?
We will use the `ChickWeight` data. Use the help to read more about the data. 
```{r}
# First look at the data.
glimpse(ChickWeight)
head(ChickWeight)
```

### A. Count how many timepoints were measured and how many chicks were on each Diet. 
```{r}
# How can you count the timepoints, chicks, and diets, and chicks nested in diets?

ChickWeight |> count(Diet)

ChickWeight |> count(Time)

ChickWeight |> count(Chick)


ChickWeight |> 
  filter(Time == "0") |>
  count(Diet) 

ChickWeight |>
  group_by(Diet) |>
  summarise(no.chicks = n_distinct(Chick))

table(ChickWeight$Time, ChickWeight$Diet)
```


### B. Now figure out which diet leads to the heaviest chicks.
```{r}
# we can plot it to get a first view
ChickWeight |>
  ggplot(aes(x = Time, y = weight, group = Chick, color = Diet)) + 
  geom_line() +
  theme_cowplot()

ChickWeight |>
  ggplot(aes(x = Time, y = weight, group = Chick, color = Diet)) + 
  geom_line() +
  theme_cowplot() +
  facet_wrap(~ Diet)


```


# How much weight gain from each Diet
```{r}
head(ChickWeight)

data <- ChickWeight |> 
  pivot_wider(names_from = Time, names_prefix = "day_", values_from = weight)


data <- data |> mutate(weight.gain = day_18 - day_0)

data |>
  ggplot(aes(x = Diet, y = weight.gain, fill = Diet)) + 
  geom_boxplot() + 
  geom_point() +
  theme_cowplot()

mod <- aov(weight.gain ~ Diet, data = data)
summary(mod)
TukeyHSD(mod)

# But we can model all the data too!

library(lme4)
library(lmerTest)

mod1 <- lmer(weight ~ Diet * Time + (1 | Chick), data = ChickWeight) # only random intercept
mod2 <- lmer(weight ~ Diet * Time + (1 + Time | Chick), data = ChickWeight) # random slope and intercept
mod3 <- lmer(weight ~ Diet * Time + (0 + Time | Chick), data = ChickWeight) # random slope only
anova(mod3, mod2)
summary(mod2)
```





