---
title: "Dimension Reduction using tidymodels"
author: "Lindsay N. Hayes"
date: 2025-07-14
draft: false
editor_options: 
  chunk_output_type: console
---

## About the activity

1)  Access the Quarto document [here](https://github.com/lindsaynhayes/introR2025/blob/main/activities/classwork05/PCA_Penguins.qmd).

2)  Download the raw file.

3) Open it in RStudio.

We will work our way through this quarto document together during class. The activity will cover Principal Component Analysis (PCA) and UMAP using tidymodels.


## Load the Packages
```{r}
#| message: false
#| error: false

library(tidyverse)
library(tidymodels)
library(palmerpenguins)
```

## `glimpse` at the data
```{r}
glimpse(penguins)
```

## Prep the Data and Perform the PCA
```{r}

# set a seed. Not really necessary for PCA because it is a reproducible analysis and doesn't use random data to model the data. But it is good practice. You can use any number your heart desires. 

set.seed(956)

# don't need to split data for the PCA because its an exploratory data analysis and we want to use all the data. 


# Build a Recipe: this is an instruction manual does not perform any actual analysis, yet. 

penguin_recipe <-
  
  # we want to use all the data to build the analysis so we us ~  and . to include all data numeric variables
  recipe( ~ ., data = penguins) |> 
  
  # we want to maintain the categorical data so we give it a new role of id to identify the data
  update_role(species, island, sex, new_role = "id") |> 
  
  # we want to omit rows with missing data because it can't be properly used by the PCA
  step_naomit(all_predictors()) |> 
  
  # to make sure all the data is normally distributed and centered we want to normalize the data before running PCA
  step_normalize(all_predictors()) |>
  
  # finally we want to actually run a PCA analysis on all the data or predictors
  step_pca(all_predictors(), id = "pca")

penguin_recipe


# Prep Data: pull in the variables from the penguins data and runs PCA

penguin_prep <- prep(penguin_recipe, penguins)

penguin_prep

penguin_prep |> tidy(id = "pca", type = "coef")
penguin_prep |> tidy(id = "pca", type = "variance")

# Bake Data: Extract the data from the PCA

penguin_bake <- bake(penguin_prep, new_data = NULL)

penguin_bake

```


## Evaluate the PCA
```{r}
# tidy type can be: "coef", "variance", 

# Plot the percent of variance explained by each principal component
penguin_prep |> 
  tidy(id = "pca", type = "variance") |> 
  dplyr::filter(terms == "percent variance") |> 
  ggplot(aes(x = component, y = value)) + 
  geom_col(fill = "#b6dfe2") + 
  xlim(c(0, 5)) + 
  ylab("% of total variance") +
  geom_text(aes(label=round(value,1)), vjust = 1.5)
```

```{r}
#| fig-width: 12
#| fig-height: 6

# Plot which predictors/variables are explained in each principal component
penguin_prep |>
  tidy(id = "pca", type = "coef") |> 
  ggplot(aes(abs(value), terms, fill = value > 0)) +
  geom_col() +
  facet_wrap(~component, scales = "free_y") +
  scale_fill_manual(values = c("#b6dfe2", "#0A537D")) +
  labs(
    x = "Absolute value of contribution",
    y = NULL, fill = "Pos Affil") 
```

```{r}
# Plot the PCA
pca_plot <- penguin_bake |> 
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(color = species, shape = species), 
             alpha = 0.8, 
             size = 2) +
  scale_colour_manual(values = c("darkorange","purple","cyan4")) +
  theme_bw()

pca_plot


# Add the vectors for how the variables are explained by each PC

pca_wider <- penguin_prep |> 
              tidy(id = "pca", type = "coef") |>
              pivot_wider(names_from = component, id_cols = terms)

arrow_style <- arrow(length = unit(.05, "inches"), type = "closed")

pca_plot +
  geom_segment(data = pca_wider,
               aes(xend = PC1, yend = PC2), 
               x = 0, 
               y = 0, 
               arrow = arrow_style) + 
  geom_text(data = pca_wider,
            aes(x = PC1, y = PC2, label = terms), 
            hjust = 0, 
            vjust = 1,
            size = 5, 
            color = '#0A537D') 

```
#### We can see that year is having a large impact of PC2 and obscuring the effect of the other metrics. So we can exclude PC2 and plot PC1 and PC3 instead. 

```{r}

pca_plot <- penguin_bake |> 
  ggplot(aes(PC1, PC3)) +
  geom_point(aes(color = species, shape = species), 
             alpha = 0.8, 
             size = 2) +
  scale_colour_manual(values = c("darkorange","purple","cyan4")) +
  theme_bw()

pca_plot

arrow_style <- arrow(length = unit(.05, "inches"), type = "closed")

pca_plot +
  geom_segment(data = pca_wider,
               aes(xend = PC1, yend = PC3), 
               x = 0, 
               y = 0, 
               arrow = arrow_style) + 
  geom_text(data = pca_wider,
            aes(x = PC1, y = PC3, label = terms), 
            hjust = 0, 
            vjust = 1,
            size = 5, 
            color = '#0A537D') 
```


## Prep the Data and Perform the UMAP instead
```{r}
library(embed)
set.seed(956)

glimpse(penguins)

penguin_recipe <-
  recipe(species ~ ., data = penguins) |> 
  update_role(species, island, sex, new_role = "id") |> 
  step_rm(year) |>
  step_naomit(all_predictors()) |> 
  step_normalize(all_predictors()) |>
  step_umap(all_numeric_predictors(), num_comp = 3)

penguin_recipe


# Prep Data: pull in the variables from the penguins data and runs PCA

penguin_prep <- prep(penguin_recipe, penguins)

penguin_prep


# Bake Data: Extract the data from the UMAP

penguin_bake <- bake(penguin_prep, new_data = NULL)

penguin_bake

```

```{r}
penguin_bake |> ggplot(aes(x = UMAP1, y = UMAP2, color = species)) +
  geom_point(alpha = .5)

penguin_bake |> ggplot(aes(x = UMAP1, y = UMAP3, color = species)) +
  geom_point(alpha = .5)

penguin_bake |> ggplot(aes(x = UMAP2, y = UMAP3, color = species)) +
  geom_point(alpha = .5)


# lets go back and try it without year and see how it looks? add in a step to the recipe: step_rm(year)
```

```{r}
sessionInfo()
```
