---
title: "Navigating Data Types in R - Answers"
author: "Lindsay N. Hayes"
date: 2025-07-03
draft: true
editor_options: 
  chunk_output_type: console
---

## About the activity

1)  Create a new Quarto document and erase the base text. You can keep the YAML header and rename it for yourself.

2)  Access the Quarto document [here](https://github.com/lindsaynhayes/introR2025/blob/main/activities/classwork02/classwork_L02.qmd) from the [github repository for the course](https://github.com/lindsaynhayes/introR2025).

3)  Copy and paste the content into your new Quarto Document. Make sure the code chunks copied correctly.

We will work our way through this quarto document together during class. The activity will cover using R as a calculator, creating R objects, and exploring the features of a data set.

## First Load the Tidyverse Package

```{r}
#| label: packages

library(tidyverse)
```

## Ways to Use R

### 1. Arithmetic

```{r}
# example: addition/subtraction/multiplication/division

193 + 45
2050 - 2025
50/250 * 100

# activity: assign the variable my_age as the current year minus your year of birth. 
```

### 2. Create R objects

```{r}
# vector
# c(..., recursive = FALSE, use.names = TRUE)
x <- c(1:225)
class(x)

# matrix
# matrix(data, nrow, ncol, byrow, dimnames)
y <- matrix(1:225, nrow=15, ncol=15, byrow = FALSE)
class(y)

# logical
# testing each variable in the vector and outputting TRUE or FALSE
over100 <- x>100
table(over100)
```

### Your Turn

### A. Create a vector of all the `homeworlds` in starwars using the `starwars` data.

```{r}

# create vector called "homeworlds" and assign it the value "homeworld" from the starwars data set
homeworlds <- starwars$homeworld

# how many worlds are there? hint: use the unique() function
unique(homeworlds)

# is there a world called "Ohio"? how would you test this with code?
table(homeworlds == "Ohio")

# How many characters live on Naboo?
table(homeworlds == "Naboo")

starwars |> filter(homeworld == "Naboo") |> count(homeworld, .drop = FALSE)

# Who lives on Naboo? (hint use the "name" variable in the starwars data and the "which" function)
unique(starwars$name[homeworlds == "Naboo"])
which(homeworlds == "Naboo")
starwars$name[which(homeworlds == "Naboo")]
```

### B. Import the `taylor` package and explore the `taylor_album_songs` dataframe

```{r}

# Try loading in the taylor package and viewing the taylor_album_songs dataframe
library(taylor)
taylor_album_songs

# what is the "class" of the object taylor_album_songs?
class(taylor_album_songs)

# what types of data are in the object taylor_album_songs?
glimpse(taylor_album_songs)
as.data.frame(map_chr(taylor_album_songs, class))

# change the "key_mode" from class "character" to class "factor"
taylor_album_songs$key_mode <- factor(taylor_album_songs$key_mode)

# How many albums are in the data set & how many songs on each album?
taylor_album_songs |> count(album_name)
```

### C. Which numeric song features are correlated with one another? Hint create a correlation matrix.

```{r}
# pick what features of the data you want to explore
colnames(taylor_album_songs)

# create a matrix of those features
mat <- taylor_album_songs[,c(1,10,11,12,13,14,15,17,18,19,20,21,22)]
colnames(mat)

# evaluate the correlation matrix
mat |> GGally::ggpairs()

# Which values show a positive correlation? Which values show a negative correlation?

#var1 <-
#var2 <-
#var3 <- 
#var4 <- 


# Plot the correlations

#ggplot(taylor, aes(x=var1, y=var2)) + geom_point(size=3)
#ggplot(taylor, aes(x=var3, y=var4)) + geom_point(size=3)

# plot loudness vs energy
ggplot(taylor_album_songs, aes(x=loudness, y=energy)) + 
  geom_point(size=3)

# plot acousticness vs energy
ggplot(taylor_album_songs, aes(x=acousticness, y=energy)) + 
  geom_point(size=3)

# plot loudness vs energy, color by album, facet by album
ggplot(taylor_album_songs, aes(x=loudness, y=energy, color = album_name)) + 
  geom_point(size=3) + 
  scale_color_albums() + 
  facet_wrap(~album_name)

# plot acousticness vs energy, color by album, facet by album
ggplot(taylor_album_songs, aes(x=acousticness, y=energy, color = album_name)) +
  geom_point(size=3) + 
  scale_color_albums() + 
  facet_wrap(~album_name)

```

