---
title: "Data Wrangling with tidyverse - Answers"
author: "Lindsay N. Hayes"
date: 2025-07-08
draft: true
editor_options: 
  chunk_output_type: console
---


## About the activity

1)  Access the Quarto document [here](https://github.com/lindsaynhayes/introR2025/blob/main/activities/classwork03/classwork_L03.qmd).

2)  Download the raw file.

3) Open it in RStudio.

We will work our way through this quarto document together during class. The activity will cover using R as a calculator, creating R objects, and exploring the features of a data set.

## Load the Tidyverse Package

```{r}
#| label: packages

library(tidyverse)
library(cowplot)
```


## Reshaping and Summarizing Data
A common type of data that requires reshaping is time course data. 

Using tidyverse principles answer the questions below:

**1. Which month had the most and least passengers in the `AirPassengers` data?**
The `AirPassengers` data which is a time-series of data representing the monthly international airline passenger numbers from January 1949 to December 1960. Search for AirPassengers in the Help to learn more about the dataset. 

```{r}
# Load and inspect the data, a little reshaping to get in to an easy to read format for you.
AP_matrix <- matrix(AirPassengers, nrow = length(unique(floor(time(AirPassengers)))), byrow = TRUE)
colnames(AP_matrix) <- month.abb
rownames(AP_matrix) <- unique(floor(time(AirPassengers)))
AP_df <- as.data.frame(AP_matrix)
AP_df$Year <- rownames(AP_matrix)
dim(AP_df)

# A. Is the data long or wide? What form does it need to be in?

dat <- AP_df |>
  pivot_longer(cols = 1:12, names_to = "month", values_to = "count")

glimpse(dat)
dat$month <- factor(dat$month, levels = month.abb)


# B. How can we extract the the most and least traveled months of the year? 

dat |>
  group_by(Year) |>
  summarise(max = max(count),
            max_month = month[which.max(count)],
            min = min(count),
            min_month = month[which.min(count)],)

dat |>
  group_by(Year) |>
  ggplot(aes (x = month, y = count, group = Year, color = Year)) + geom_line()
```

**2. What was the percent increase in passengers each year between Aug and Nov?**
```{r}
# To answer this question we need to find the ratio of Aug and Nov travelers. We need the data in the wide format.

AP_df

# how can we add the ratio?
AP_df |> mutate(ratio = ((Aug/Nov)-1)*100) |>
  ggplot(aes(x = Year, y = ratio)) + geom_point(size = 3)

```


3. Which diet caused the chicks to grow more?
We will use the `ChickWeight` data. Use the help to read more about the data. 
```{r}
# First lets just look at the data.
glimpse(ChickWeight)

# Second, look at the distribution of the data. Count how many timepoints was measured nad how many chicks were in each Diet. 

ChickWeight |> count(Diet)
ChickWeight |> count(Chick)
n_distinct(ChickWeight$Chick)

ChickWeight |> count(Time)


ChickWeight |> 
  group_by(Diet) |>
  count(Chick)

ChickWeight |> 
  filter(Time == 0) |>
  count(Diet)

ChickWeight |> 
  group_by(Diet) |>
  summarize(no.chicks = n_distinct(Chick))

```
Ok so we know there are 4 diets with 20 chicks on Diet 1, and 10 chicks each on diets 2-4.

Now lets figure out which diet leads to the heaviest chicks.
```{r}
# we can plot it to get a first view

ChickWeight |>
  ggplot(aes(x = Time, y = weight, group = Chick, color = Diet)) + 
  geom_line() +
  theme_cowplot()

ChickWeight |>
  pivot_wider(names_from = Time, names_prefix = "day_", values_from = weight) |>
  mutate(weight.gain = day_18 - day_0) |>
  ggplot(aes(x = Diet, y = weight.gain, fill = Diet)) + geom_boxplot()

data <- ChickWeight |>
  pivot_wider(names_from = Time, names_prefix = "day_", values_from = weight) |>
  mutate(weight.gain = day_18 - day_0)

mod <- aov(weight.gain ~ Diet, data = data)
summary(mod)
TukeyHSD(mod)
pairwise.t.test(data$weight.gain, data$Diet, p.adjust.method = "BH")

library(lme4)
library(lmerTest)

mod <- lmer(weight ~ Diet * Time + (1|Chick), data = ChickWeight)
summary(mod)

```
