---
title: "Data Visualization with ggplot2"
author: "Lindsay N. Hayes"
date: 2025-07-08
draft: true
editor_options: 
  chunk_output_type: console
---

## About the activity

1)  Access the Quarto document [here](https://github.com/lindsaynhayes/introR2025/blob/main/activities/classwork04/classwork_L04.qmd).

2)  Download the raw file.

3) Open it in RStudio.

We will work our way through this quarto document together during class. The activity will cover plotting data with ggplot2!

## Load the Packages
The `tidyverse` library includes `ggplot2` and I added `cowplot` to make the plots prettier. We will use data from the `palmerpenguins` and the `taylor` packages.
```{r}
#| label: packages
#| message: false

library(tidyverse)
library(cowplot)
library(palmerpenguins)
library(taylor)
```

## Histogram & Density plots
Lets use the `palmerpenguins` data to plot some histograms. Histograms are good for evaluating the distribution of continuous data.

#### 1. How would you describe the distribution? Is it uni-modal or bi-modal?
```{r}

penguins |> head()

# plot the distribution of the penguins' body mass using geom_histogram()

ggplot(penguins, aes(x = body_mass_g)) + geom_histogram()

# lets make it a little more pretty (note tidy syntax is to put each element on a new line, the "+" wraps the text)

ggplot(penguins, aes(x = body_mass_g)) + 
  geom_histogram(fill = "deeppink") + 
  theme_cowplot()

# histograms are impacted by the "bin" or how each bar is averaging (how thick the bars are). We can change that will `bin` or `binwidth`

ggplot(penguins, aes(x = body_mass_g)) + 
  geom_histogram(binwidth = 25, fill = "deeppink") + 
  theme_cowplot()

# add some more labels

ggplot(penguins, aes(x = body_mass_g)) + 
  geom_histogram(binwidth = 50, fill = "deeppink") + 
  theme_cowplot() +
  labs(title = "Distribution of Penguin Body Mass",
         x = "Body Mass (g)",
         y ="Count")

```

#### 2. Plot the other penguin bill and flipper measurements? What are their distribution? How does the binwidth change the interpretation of the distributions?
```{r}

penguins |> head()

# try 3
ggplot(penguins, aes(x = bill_length_mm)) + 
  geom_histogram(binwidth = 1, fill = "purple") + 
  theme_cowplot() +
  labs(title = "Distribution of Penguin Bill Length",
         x = "Bill Length (mm)",
         y ="Count")


ggplot(penguins, aes(x = bill_depth_mm)) + 
  geom_histogram(binwidth = 0.3, fill = "springgreen") + 
  theme_cowplot() +
  labs(title = "Distribution of Penguin Bill Depth",
         x = "Bill Depth (mm)",
         y ="Count")

ggplot(penguins, aes(x = flipper_length_mm)) + 
  geom_histogram(binwidth = 1, fill = "gold") + 
  theme_cowplot() +
  labs(title = "Distribution of Penguin Flipper Length",
         x = "Flipper Length (mm)",
         y ="Count")

```

#### 3. A benefit of geom_density() is that the plots are not dependent on the binsize. Compare the histograms and the density plots for each data feature.
```{r}

a <- ggplot(penguins, aes(x = body_mass_g)) + 
  geom_density(fill = "deeppink") + 
  theme_cowplot() +
  labs(subtitle = "Distribution of Penguin Body Mass",
         x = "Body Mass (g)",
         y ="Density")

b <- ggplot(penguins, aes(x = bill_length_mm)) + 
  geom_density(fill = "purple") + 
  theme_cowplot() +
  labs(subtitle = "Distribution of Penguin Bill Length",
         x = "Bill Length (mm)",
         y ="Density")


c <- ggplot(penguins, aes(x = bill_depth_mm)) + 
  geom_density(fill = "springgreen") + 
  theme_cowplot() +
  labs(subtitle = "Distribution of Penguin Bill Depth",
         x = "Bill Depth (mm)",
         y ="Density")

d <- ggplot(penguins, aes(x = flipper_length_mm)) + 
  geom_density(fill = "gold") + 
  theme_cowplot() +
  labs(subtitle = "Distribution of Penguin Flipper Length",
         x = "Flipper Length (mm)",
         y ="Density")

plot_grid(a,b,c,d, labels = c("A", "B", "C", "D"))

```

#### 4. Where do you think the bimodal distributions may come from? Lets check the distribution across different species. Add `fill = species` to the `aes()` so we can see how the distributions change across species of penguins. 
```{r}
#| fig-width: 10
#| fig-height: 6
#| warning: false
 

# add an `alpha` to the geom_density to sets the opacity of the density plots. 

a <- ggplot(penguins, aes(x = body_mass_g, fill = species)) + 
  geom_density(alpha = 0.5) + 
  theme_cowplot() +
  labs(subtitle = "Distribution of Penguin Body Mass",
         x = "Body Mass (g)",
         y ="Density")

b <- ggplot(penguins, aes(x = bill_length_mm, fill = species)) + 
  geom_density(alpha = 0.5) + 
  theme_cowplot() +
  labs(subtitle = "Distribution of Penguin Bill Length",
         x = "Bill Length (mm)",
         y ="Density")


c <- ggplot(penguins, aes(x = bill_depth_mm, fill = species)) + 
  geom_density(alpha = 0.5) + 
  theme_cowplot() +
  labs(subtitle = "Distribution of Penguin Bill Depth",
         x = "Bill Depth (mm)",
         y ="Density")

d <- ggplot(penguins, aes(x = flipper_length_mm, fill = species)) + 
  geom_density(alpha = 0.5) + 
  theme_cowplot() +
  labs(subtitle = "Distribution of Penguin Flipper Length",
         x = "Flipper Length (mm)",
         y ="Density")

plot_grid(a,b,c,d, labels = c("A", "B", "C", "D"))
```




## Barplots of Categorical data
Another type of distribution and count data is of categorical data. Count data for categorical data is best displayed in barplots

```{r}
#| fig-width: 8
#| fig-height: 6

# some tidying
taylor_album_songs$key <- str_sub(taylor_album_songs$key_name, 1, 1)
taylor_album_songs$album_name <- fct_inorder(taylor_album_songs$album_name)

# You can see both the difference in the number of songs per album, but also in the numbers in each key. 

ggplot(taylor_album_songs, aes(y = album_name, fill = key)) + 
  geom_bar() +
  theme_cowplot() +
  labs(title = "Keys of TS albums",
       x = "Count", 
       y = "") + scale_fill_brewer(palette="Set2")

# Now lest look at the proportions
ggplot(taylor_album_songs, aes(y = album_name, fill = key)) + 
  geom_bar(position = "fill") +
  theme_cowplot() +
  labs(title = "Keys of TS albums",
       x = "Proportion", 
       y = "") + scale_fill_brewer(palette="Set2")

```

#### 5. What factors distinguish Ideal cut versus Good cut diamonds? 
Using the `diamonds` dataset, determine the distribution of the `cut`, `color`, and `clarity` of diamonds.
```{r}
#| fig-width: 10
#| fig-height: 7

a <- ggplot(diamonds, aes(y = cut)) + 
  geom_bar(fill = "yellow", color = "black") +
  theme_cowplot() 

b<- ggplot(diamonds, aes(x = color)) + 
  geom_bar(fill = c(rainbow(7))) +
  theme_cowplot()

c <- ggplot(diamonds, aes(y = clarity, fill = clarity)) + 
  geom_bar() +
  theme_cowplot() +
  scale_fill_grey()

plot_grid(a,b,c)

# Which plots show the effect best?
a1 <- ggplot(diamonds, aes(y = color, fill = cut)) + 
  geom_bar(position = "fill") +
  theme_cowplot() 


a2 <- ggplot(diamonds, aes(y = clarity, fill = cut)) + 
  geom_bar(position = "fill") +
  theme_cowplot() 

a3 <- ggplot(diamonds, aes(y = cut, fill = color)) + 
  geom_bar(position = "fill") +
  theme_cowplot() 


a4 <- ggplot(diamonds, aes(y = cut, fill = clarity)) + 
  geom_bar(position = "fill") +
  theme_cowplot() 



plot_grid(a1, a2, a3, a4)
```


## Boxplots! 
The best way to view group differences.

#### 6. Does the bill length and depth, the flipper length, or the body mass differ across species of penguins? 

Using the `penguins` dataset use geom_boxplot() to find out if there is a difference in the penguin features between different species of penguins 

```{r}
#| fig-width: 8
#| fig-height: 6

# quick plot
a <- ggplot(penguins, aes(x = species, y = bill_length_mm, fill = species)) + geom_boxplot()
b <- ggplot(penguins, aes(x = species, y = bill_depth_mm, fill = species)) + geom_boxplot()
c <- ggplot(penguins, aes(x = species, y = flipper_length_mm, fill = species)) + geom_boxplot()
d <- ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) + geom_boxplot()

plot_grid(a,b,c,d)
```
```{r}
# make a nice plot
ggplot(penguins, aes(x = species, y = bill_length_mm, fill = species)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(size = 2, aes(shape = sex), position = position_jitter(width = 0.1)) +
  theme_cowplot() +
  ylim(30,60) +
  labs(y = "Bill Length (mm)", x = "", subtitle = "Bill Length across Species")


ggplot(penguins, aes(x = species, y = bill_depth_mm, fill = species)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(size = 2, aes(shape = sex), position = position_jitter(width = 0.1)) +
  theme_cowplot() +
  ylim(10,25) +
  labs(y = "Bill Depth (mm)", x = "", subtitle = "Bill Depth across Species")


ggplot(penguins, aes(x = species, y = flipper_length_mm, fill = species)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(size = 2, aes(shape = sex), position = position_jitter(width = 0.1)) +
  theme_cowplot() +
  ylim(150,250) +
  labs(y = "Flipper Length (mm)", x = "", subtitle = "Flipper Length across Species")


ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_point(size = 2, aes(shape = sex), position = position_jitter(width = 0.1)) +
  theme_cowplot() +
  ylim(2500,7000) +
  labs(y = "Mass (g)", x = "", subtitle = "Mass across Species")

```

*Isn't this better than bar graphs??!!!*

**Note: `geom_violin()` can also be a good alternative to a boxplot if you have a very large sample size, such as in single cell RNA sequencing or flow cytometry data. Try the plots as a `geom_violin()` instead of `geom_boxplot()`.**


## Scatter and Line Plots 
Looking at 2 continuous variables and their relationship to one another. 

#### 7. Using the `iris` dataset use to find out the relationship between `Sepal.Length`, `Sepal.Width`, `Petal.Length`, and Petal.Width` and how that relationship changes across different `Species` of iris flowers. 


```{r}

iris |> head()


ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species, group = Species)) + 
  geom_point() + 
  geom_smooth(method = lm) +
  theme_cowplot() +
  labs( x = "Sepal Length (mm)", y= "Sepal Width (mm)")



ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species, group = Species)) + 
  geom_point() + 
  geom_smooth(method = lm) +
  theme_cowplot() +
  labs( x = "Petal Length (mm)", y= "Petal Width (mm)")


```


#### 8. Now use the `ChickWeight` data to determine which `Diet` lead to the greatest `weight` gain over `Time.` Remember to take into account repeated measures across `Chick`


```{r}




```


```{r}
sessionInfo()
```

